A Way Forward in Parallelising Dynamic Languages

### 2014-05-05 talk/icooolps2014/position-paper.tex
change title
a few typos

\title{The Way Forward in Parallelising Dynamic Languages}
\subtitle{Position Paper, ICOOOLPS'14}

@@

\begin{abstract}
  Dynamic languages became very popular in recent years. At some
  point, the need for concurrency arised, and many of them made the
  choice to use a single global interpreter lock (GIL) to synchronize
  the interpreter in a multithreading scenario. This choice however
  makes it impossible to actually run code in parallel.

  Here we want to compare different approaches to replacing the GIL
  with a technology that allows parallel execution. We look at
  fine-grained locking, shared-nothing, and transactional memory (TM)
  approaches. We argue that software-based TM systems are the most
  promising, especially since they also enable the introduction of
  atomic blocks as a better synchronization mechanism in the language.
\end{abstract}

@@

%% Furthermore, a solution to this problem should also bring better
%% synchronization mechanism with it...

%% (supporting (large) atomic blocks for synchronization)

@@

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in the interpreter. The basic guarantee is that the GIL
may only be released in-between bytecode instructions. The interpreter
can thus rely on complete isolation and atomicity of these
instructions. Additionally, it provides the application with a
sequential consistency model\cite{lamport79}. As a consequence,
applications can rely on certain operations to be atomic and that they
will always be executed in the order in which they appear in the
code. While depending on this may not always be a good idea, it is
done in practice. A GIL-replacement should therefore uphold these
guarantees, while preferably also be as easily implementable as a GIL
for the interpreter. The latter can be especially important since
many of these languages are developed and maintained by very large
open-source communities, which are not easy to coordinate.

@@

Unfortunately, STM has a big performance problem. Particularly, for
our use case there is not much static information available since we
are executing a program only known at runtime. Additionally, replacing
the GIL means running every part of the application in transactions,
so there is not much code that can run outside and that can be
optimized better. The performance of the TM system is vital.

>>>>

\title{A Way Forward in Parallelising Dynamic Languages}
\subtitle{Position Paper, ICOOOLPS'14}

@@

\begin{abstract}
  Dynamic languages became very popular in recent years. At some
  point, the need for concurrency arose, and many of them made the
  choice to use a single global interpreter lock (GIL) to synchronise
  the interpreter in a multithreading scenario. This choice however
  makes it impossible to actually run code in parallel.

  Here we want to compare different approaches to replacing the GIL
  with a technology that allows parallel execution. We look at
  fine-grained locking, shared-nothing, and transactional memory (TM)
  approaches. We argue that software-based TM systems are the most
  promising, especially since they also enable the introduction of
  atomic blocks as a better synchronisation mechanism in the language.
\end{abstract}

@@

%% Furthermore, a solution to this problem should also bring better
%% synchronisation mechanism with it...

%% (supporting (large) atomic blocks for synchronisation)

@@

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions. The interpreter
can thus rely on complete isolation and atomicity of these
instructions. Additionally, it provides the application with a
sequential consistency model\cite{lamport79}. As a consequence,
applications can rely on certain operations to be atomic and that they
will always be executed in the order in which they appear in the
code. While depending on this may not always be a good idea, it is
done in practice. A GIL-replacement should therefore uphold these
guarantees, while preferably also be as easily implementable as a GIL
for the interpreter. The latter can be especially important since
many of these languages are developed and maintained by very large
open-source communities, which are not easy to coordinate.

@@

Unfortunately, STM has a big performance problem. Particularly, for
our use case there is not much static information available since we
are executing a program only known at runtime. Additionally, replacing
the GIL means running every part of the application in transactions,
so there is not much code that can run outside and that can be
optimised better. The performance of the TM system is vital.
